---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R tools for spatial data at the Australian Antarctic Division (AAD)

Tools for reading, plotting and manipulating spatial data used at the Australian Antarctic Division (AAD).

The typical use-cases for raadtools are

* read a time series gridded data set as a function of date, optionally with spatial subsetting
* match a data set of longitude, latitude, time to the corresponding value in a time series gridded data set

Examples of these workflows are outlined in this [rOpenSci blog post](https://ropensci.org/blog/2018/11/13/antarctic/). 

There are oceanography, topography, meteorology, altimetry, sea ice, ocean colour, and many other data sources. These are mostly remote sensing but including re-analysis and model output products. 

The package uses the R [raster package]( https://CRAN.R-project.org/package=raster) and always provides data as a standard raster (RasterLayer, RasterBrick, or RasterStack). Each data set is invdividually handled by a function to ensure the spatial and temporal registration is correct. There are several `read*()` functions in raadtools, see the full list in the documentation: 

http://australianantarcticdivision.github.io/raadtools/reference/index.html

To build the underlying data collection raadtools relies on the [bowerbird](https://github.com/ropensci/bowerbird) and [blueant](https://github.com/AustralianAntarcticDivision/blueant) packages. 


# Acess to raadtools

Typically you will be provided with access, and won't be aware of the underlying details, but the repository of data used by raadtools is available under RDSI/PUBLIC/raad and at the AAD under gridded_new/ in the scientific data collection.  

Anyone with a [Nectar account](https://dashboard.rc.nectar.org.au) may run this by creating a VM from our `raadclient` image. Search the public images for **raadclient** (e.g. '	raadclient06_20181016' but choose the latest one) and ensure that the SSH and RStudio port (8787) is open. Use the default *rstudio/rstudio* account, or create your own. 

The [contents of the repository](https://github.com/AustralianAntarcticDivision/blueant#data-source-summary) is listed in the
technical configuration for bowerbird. 

If you would like a collection added please make a request via a [Github issue](https://github.com/AustralianAntarcticDivision/bowerbird/issues/new) or contact one of the authors directly. 

You are welcome to make your own copies of data from the collection for your own use, but please respect the citation and usage requests of the data providers [listed in the summary](https://github.com/AustralianAntarcticDivision/blueant#data-source-summary).


# Using raadtools 

There are two main ways to use it. 

## 1. RStudio raadtools server

If you have access to a "raadtools-RStudio-server" then you need only load the package to get started: 

```{r,eval=FALSE}
library(raadtools)
```

## 2. Local computer, within the AAD network

If it's not installed, trying installing with 

```{r,eval=FALSE}
devtools::install_github("AustralianAntarcticDivision/raadtools")
library(raadtools)
```

If neither of 1 or 2 work for you you, see your local raadtools expert. 



Please note that the 'raadtools' project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
