---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Travis-CI Build Status](https://travis-ci.org/AustralianAntarcticDivision/raadtools.svg?branch=master)](https://travis-ci.org/AustralianAntarcticDivision/raadtools)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/AustralianAntarcticDivision/raadtools?branch=master&svg=true)](https://ci.appveyor.com/project/AustralianAntarcticDivision/raadtools)
[![Coverage Status](https://img.shields.io/codecov/c/github/AustralianAntarcticDivision/raadtools/master.svg)](https://codecov.io/github/AustralianAntarcticDivision/raadtools?branch=master)


# R tools for spatial data at the Australian Antarctic Division (AAD)

Tools for reading, plotting and manipulating spatial data used at the Australian Antarctic Division (AAD).

There are two main ways to use it. 

## 1. RStudio raadtools server

If you have access to a "raadtools-RStudio-server" then you need only load the package to get started: 

```{r,eval=FALSE}
library(raadtools)
```

## 2. Local computer, within the AAD network

If it's not installed, trying installing with 

```{r,eval=FALSE}
devtools::install_github("AustralianAntarcticDivision/raadtools")
library(raadtools)
```

If neither of 1 or 2 work for you you, see your local administrator. 


## Development

raadsync keeps it up to date:  

https://github.com/AustralianAntarcticDataCentre/raadsync

A configuration like this is required for the tools to work: 

https://github.com/mdsumner/nectar/blob/master/r-spatial.sh


## Related, or similar, projects

https://github.com/mdsumner/roc - read ocean colour files catalogued by raadtools::ocfiles()

https://github.com/ropensci/rerddap

https://ropensci.org/tutorials/rnoaa_tutorial.html

https://github.com/rmendels/xtractomatic



