---
title: "Remotely Sensed Ocean Colour (Chlorophyll) in raadtools"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Remotely Sensed Ocean Colour (Chlorophyll) in raadtools}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(raadtools)
```

The function `readchla` can be used to read a mean value of remotely sensed chlorophyll-a for an arbitrary set of dates (days) within a region, either from the MODISA (2002-) or SeaWiFS (1997-2012) platform and with the Johnson 2013 or NASA algorithm. 

For example, provide a vector of dates, and a spatial extent (longlat) to produce a raster map at the native resolution of the Level-3 Mapped product for given product. 


Note that combining different products and longer time periods can help augment the missing coverage spatially, but ultimately there are cloud issues and so spatial averaging and statistics is required for more comprehensive data coverage. 

```{r}
library(raadtools)
dates <- seq(as.Date("2020-01-01"), by = "1 day", length = 6)
ext <- extent(140, 150, -60, -40)
modis <- readchla(dates, ext, product = "MODISA")
#seawifs <- readchla(dates, ext, product = "SeaWiFS")  ## no SW in the collection atm
viirs <- readchla(dates, ext, product = "VIIRS")
op <- par(mfrow = c(1, 3))
chlp <- palr::chl_pal(palette = TRUE)
plot(modis, col =chlp$cols, breaks = chlp$breaks, legend = FALSE)
plot(viirs, col =chlp$cols, breaks = chlp$breaks, legend = FALSE)
plot(mean(stack(modis, viirs), na.rm = TRUE), col =chlp$cols, breaks = chlp$breaks, legend = FALSE)
par(op)


```


