---
title: "Remotely Sensed Ocean Colour (Chlorophyll) in raadtools"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Remotely Sensed Ocean Colour (Chlorophyll) in raadtools}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(raadtools)
```

The function `readchla` reads remotely sensed chlorophyll-a for an arbitrary set of dates (days) within a region, from the MODISA (2002-), SeaWiFS (1997-2012) or VIIRS platform, with the NASA algorithm. 

For example, provide a vector of dates, and a spatial extent (longlat) to produce a terra `SpatRaster` at the native resolution of the Level-3 Mapped product, with one layer per date. To get a single map, average the layers with `mean()`. (Earlier versions of `readchla` did that averaging for you and returned one layer.)


Note that combining different products and longer time periods can help augment the missing coverage spatially, but ultimately there are cloud issues and so spatial averaging and statistics is required for more comprehensive data coverage. 

```{r}
library(raadtools)
dates <- seq(as.Date("2020-01-01"), by = "1 day", length = 6)
roi <- ext(140, 150, -60, -40)
modis <- readchla(dates, xylim = roi, product = "MODISA")
#seawifs <- readchla(dates, xylim = roi, product = "SeaWiFS")  ## no SW in the collection atm
viirs <- readchla(dates, xylim = roi, product = "VIIRS")
modis
op <- par(mfrow = c(1, 3))
chlp <- palr::chl_pal(palette = TRUE)
plot(mean(modis, na.rm = TRUE), col = chlp$cols, breaks = chlp$breaks, legend = FALSE)
plot(mean(viirs, na.rm = TRUE), col = chlp$cols, breaks = chlp$breaks, legend = FALSE)
plot(mean(c(modis, viirs), na.rm = TRUE), col = chlp$cols, breaks = chlp$breaks, legend = FALSE)
par(op)


```


