---
title: "Remotely Sensed Ocean Colour (Chlorophyll) in raadtools"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chlorophyll}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The function `readchla` can be used to read a mean value of remotely sensed chlorophyll-a for an arbitrary set of dates (days) within a region, either from the MODISA (2002-) or SeaWiFS (1997-2012) platform and with the Johnson 2013 or NASA algorithm. 

For example, provide a vector of dates, and a spatial extent (longlat) to produce a raster map at the native resolution of the Level-3 Mapped product for given product. 

```{r}
library(raadtools)
dates <- seq(as.Date("2010-01-01"), by = "1 day", length = 20)
ext <- extent(140, 150, -60, -40)
modis <- readchla(dates, ext, product = "MODISA")
seawifs <- readchla(dates, ext, product = "SeaWiFS")

op <- par(mfrow = c(1, 2))
chlp <- palr::chlPal(palette = TRUE)
plot(modis, col =chlp$cols, breaks = chlp$breaks, legend = FALSE)
plot(seawifs, col =chlp$cols, breaks = chlp$breaks, legend = FALSE)
par(op)


```

There is a fairly complex stack of infrastructure underneath this function to make this possible. 

The L3 RRS bin data is downloaded and processed for Johnson and NASA chlorophyll,  and stored in sparse form as a table with bin number and chlorophyll-a values. 




