---
title: "Environmental data from an extent with raadtools"
author: "Michael Sumner and Jaimie Cleeland"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 9
    fig_height: 9
vignette: >
  %\VignetteIndexEntry{Environmental data from an extent with raadtools}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r,include=FALSE}

library(raadtools)
```
The `raadtools` package contains a number of `read[something]` functions, mostly for gridded time-series remote sensing data. 

All of the read functions have an "xylim" argument, which will take a terra `ext(xmin, xmax, ymin, ymax)`. E.g. 

```{r}

library(raadtools)
library(raadfiles)
read_oisst_daily("2010-01-01", xylim = ext(100, 160, -50, -40))
```

That extent must make sense for the data set though, so see what this returns first (it will be a longlat-grid in [-180, 180, -90, 90]. 


Note that we ran the function above *without saving the result anywhere*, so we just saw a print-out summary of the result but it is now discarded. 

That is a useful technique to see what a function does, since it will return a layer by default.  This can be plotted or used by other functions, and so is the best way to learn about the data source. 

```{r}
library(raadtools)
library(raadfiles)
read_oisst_daily()
```

That works for `readssh`, `readcurr`, and `readwind` just the same, and the date argument can be a single date or a vector of them. Every read function returns a terra `SpatRaster`, which is one class whether it holds one layer or many: one layer per date, with the date-time of each layer recorded on it (see `terra::time()`). After the read function is finished we are back to using the `terra` package, so familiarity with that package and its documentation is important. If you are used to the `raster` package, the transition vignette in raadtools has a short translation table. 

```{r,eval=FALSE}
help("terra-package", package = "terra")
vignette("terra-transition", package = "raadtools")
```

Some of the functions respect `lon180 = FALSE` because they have a grid that was originally in `[0, 360, -90, 90]` and in that case it can be faster to use that domain for the extent rather than the common `[-180, 180, -90, 90]` one.   (Please feel free to ask about this for specific data sets, it's one of those boring details). 

For ice, and some other data, the grids are not in longlat so `xylim` needs an extent in that projection. You can always draw one like this, with code that must be run interactively as the plot will wait for you to click twice on the plot (defining a rectangle) before the `e` object is created. 

```{r,eval=FALSE}
ice <- read_nsidc_cdr_daily()  
plot(ice)
e <- draw()
```

```{r,eval=TRUE, echo=FALSE}
library(raadtools)
library(raadfiles)
ice <- read_nsidc_cdr_daily()  
plot(ice)
e <- ext(1059036.6077237, 3280191.02253238, 873709.267254125, 3119406.82476787)
lines(e, col = "red")
```

Notice that we can also make `SpatExtent` objects from scratch, and perform operations on them (`+ 10` moves each edge out by 10). 

```{r}
my_extent <- ext(100, 120, -40, -20)
my_extent
my_extent + 10

```


otherwise `project()` will transform a longlat extent into the grid's projection. It's not that straightforward because of the polar aspect, which is why `project()` adds points along each edge of the extent before transforming it, so the result covers the curved edges and not just the four corners. 

Sea ice concentration is a fraction from 0 to 1. The `range` below leaves open water (below 1%) uncoloured. 

```{r}
library(raadtools)
library(raadfiles)
ice <- read_nsidc_cdr_daily() 
pex <- project(ext(100, 160, -70, -50), "OGC:CRS84", crs(ice))

ice_ex <- read_nsidc_cdr_daily("2016-09-01", xylim = pex)

plot(ice_ex, col = grey(seq(0, 1, length = 100)), range = c(0.01, 1))
```


Please note that the `xylim` argument just applies `crop`, so you can avoid it altogether and do it after the fact. 

```{r}
library(raadtools)
library(raadfiles)
crop(read_nsidc_cdr_daily(),  ext(100, 160, -50, -40), snap = "out")
```

(`snap = "out"` makes the intersection *inclusive*, i.e. any overlapping part of a pixel means that pixel is included in the result. By default `crop` snaps to the nearest cell edge, which can drop partial overlaps on the edge). 


A long time series is cheap to read, because terra does not read cell values until something asks for them. To keep a cropped series, crop it with xylim and write it out to a file: 

```R
sst <- readsst(dates,  xylim = ext(100, 160, -50, -40))
writeRaster(sst, "mymegasplat.tif")
```

More details ... TBD


## Extending maps

With two examples from above. 


Plot the projected data with a graticule, and add a "transect". 

```{r}
library(raadtools)
library(raadfiles)
ice <- read_nsidc_cdr_daily() 
pex <- project(ext(100, 160, -70, -50), "OGC:CRS84", crs(ice))

ice_ex <- readice("2016-09-01", xylim = pex)

## create a graticule for this region

g <- graticule(seq(100, 160, by = 15), seq(-75, -50, by = 5), crs = crs(ice))
plot(ice_ex, col = grey(seq(0, 1, length = 100)), range = c(0.01, 1), axes = FALSE, 
     fun = function() lines(g, lty = 2, col = "firebrick"))

## add a "transect"
lons <- seq(100, 160, by = 1)
transect <- project(cbind(lons, -63), "OGC:CRS84", crs(ice_ex))


points(transect, cex = 0.5, pch = 19, col = "dodgerblue")
```

Extract values at specific points. This is terra's `extract()` on a grid we have already read, so every point takes its value from the one date in `ice_ex`. It returns a data.frame with one column per layer. 

```{r}
library(raadtools)
library(raadfiles)
iceconc <- extract(ice_ex, transect, method = "bilinear")
plot(lons, iceconc[[1]], main = "ice concentration at latitude -63 between 100E and 160E")
```

When each point has its own date, give `extract()` the read function itself rather than a grid, with a data.frame of longitude, latitude and date-time in that order. Only the dates the points need are read, and each point takes its value from the latest layer at or before its date-time (see `?raadtools::extract`, and the xyt package that does the work). 

```{r,eval=FALSE}
## optionally set daemons for parallelization
#mirai::daemons(12)  ## and unset later #mirai::daemons(0)

pts <- data.frame(lon = lons, lat = -63, 
                  date = seq(as.Date("2016-09-01"), by = "1 day", length.out = length(lons)))
extract(read_nsidc_cdr_daily, pts)
```



Plot mean SST and contour. 

```{r}
library(raadtools)
library(raadfiles)
dates <- seq(as.Date("2016-01-01"), length = 31, by = "1 day")
sst <- read_oisst_daily(dates,  xylim = ext(100, 160, -50, -40))
msst <- app(sst, mean)
sdsst <- app(sst, sd)
plot(msst, col = hcl.colors(50, "viridis"), main = "mean SST Jan 2016",  fun = function() contour(sdsst, add = TRUE))
```