%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Installing and maintaining raadtools}

# Installing the raadtools package, Michael Sumner

This document is currently relevant to working on Windows, much of the
same stuff will work on Linux but no care has been taken to point out
the differences here.

## Setup

To install the __raadtools__ package from within the AAD you
should be able to run the following code from within R. A relatively
recent version of R will be assumed, and definitely no earler than
__R 2.15.0__ can be expected to work.

This is the basic full install and configuration, see below for each step in greater
detail.

```{r eval=FALSE,tidy=FALSE}
install.packages(c("maptools", "raster", "testthat", "ncdf", "rgdal", "rgeos", "knitr"),
                 dependencies = TRUE)
install.packages("raadtools", repos ="file://aad.gov.au/files/Transfer/toolbox/RAAD")
```

Now to load the package, use the usual package loading code here. If
there is a problem with determining the path to the data repository
this will provide a message about that.

```{r eval=FALSE}
library(raadtools)
```

If a message reports that your data repository cannot be found, follow
the instructions to set it and send this package author your details.

The instructions go like this, by setting an option called
__default.datadir__, note that this only sets this for the current R
session and you will have to repeat it each time until a better
solution is developed.

```{r eval=FALSE}
options(default.datadir = "/path/to/repository/of/data")
```

(The tip of the path there "data" should include files such as "chl",
"wind", "sst", etc. i.e. it looks like data/chl, data/wind, data/sst
which hold the individual families of the main data types).

## Setup details

The remaining text in this section repeats the installation steps
above individually, this might be required for customized
installations or development.

If the following step fails you may not have your network settings
setup appropriately, so first run __setInternet2__ and then try to
install __raster__ again. (This option can be set for an R installation
by choosing "Internet2" when R is installed, and this is what you
should do. Otherwise, R is launced with "Rgui.exe" and this can be
invoked with the option to setInternet2 by using __Rgui.exe --internet2__ 
which can be set on a Windows shortcut under Properties).

```{r eval=FALSE}
setInternet2(TRUE)
install.packages(c("raster", "maptools"))
```

The following packages are not essential to run __raadtools__ but
otherwise some features will not be available. To get the full
functionality please do run the following installations as well.


```{r eval=FALSE}
install.packages("ncdf", "rgdal", "rgeos")
```

Finally, install the latest version of __raadtools__ from the
local AAD repository.

```{r eval=FALSE}
install.packages("raadtools", repos ="file://aad.gov.au/files/Transfer/toolbox/RAAD")
```

## Development of raadtools
In order to contribute to the development of raadtools, the following options are available. 

- Complain, report, plead, compel the package author/s. Polite emails
  are recommended and hardware store vouchers welcome, but do what you must to get the message through.
- Get hold of the source, make your changes and send patches to the author/s. See previous option. 
- Check out the source with Git, and push your changes through to the source repository. 

The following section expands on the Git option. 

## Process of committing changes to the Git source of raadtools

The raadtools package source is hosted on a Stash site at the AAD. This is located here, but for now we won't mention Stash again. 

[Stash at the AAD](http://stash.aad.gov.au/projects/MDS/repos/raadtools/browse)

### Install prerequisites for building raadtools

1. Install the most recent version of R, and set its location in your PATH. 
2. Install [RTools](http://cran.r-project.org/bin/windows/Rtools/) relevant to the latest version of R, and elect that it sets its location in your PATH (the installer can do this). 
3. Install [MikTeX](http://miktex.org/download) and ensure it is in your PATH (later versions do this on install). 
4. Install the full list of package dependencies listed above. 

### Clone the raadtools repository

You can get the raadtools source by cloning it with Git: 

```
git clone http://michae_sum@stash.aad.gov.au/scm/mds/raadtools.git
```

### Building the raadtools package from the Git source

The __raadtools__ package must be pre-processed with
[roxygen2](http://cran.r-project.org/web/packages/roxygen2/index.html)
before it can be built. This also requires extra packages that are not
necessary for using the package. These and their dependencies can be
installed like this

```{r eval=FALSE}
install.packages(c("testthat", "knitr", "roxygen2"))
```

To process the source package that has been checked out from Git,
start R and run the following code.  This code processes the source
__.R__ files in R/ and generates the man/*.Rd files required for the
sources (multiple R code objects may be documented as a single .Rd
file).

```{r eval=FALSE}
library(roxygen2)
roxygenize("raadtools")
```

Then, run the usual set of R commands to build source, install and
build a binary, and check the package. Note that this example code necessarily
denotes a specific version of the package and so the relevant text
will be different for the actual process when you try it.

```{shell eval=FALSE}
R CMD build raadtools
R CMD INSTALL raadtools_0.0-27.tar.gz --build
R CMD check raadtools_0.0-27.tar.gz
```

Because I work with the Git Bash shell for this, I have the following
workflow set up that simplifies a lot of this. This relies on the fact
that I have all of the the dependencies listed above. (There can be
complications with Cygwin and other unix-like tools installed on
Windows, but if ls/zip/latex/git/R all work for you at the command
line you should be good to go).

1. Open the command shell, which I have set to "Start In" the parent directory of "raadtools".  (I happen use the Git Bash shell, which I already have open for working with Git, but note that this doesn't have all of the Windows functionality avaialable to it)
2. run shell script "roxy raadtools" (see below)
3. Run R CMD build/INSTALL/check as above
4. Wash, rinse, repeat

There are tests that will run via the __testthat__ package, but note
the following options that can be used to force a quick informal
build/INSTALL/check:

```{shell eval=FALSE}
R CMD build raadtools --no-build-vignettes
R CMD INSTALL raadtools_0.0-28.tar.gz --build
R CMD check raadtools_0.0-28.tar.gz --no-tests
```

See the help page for each individual command. 

Definition of the "roxy" shell script. 
```{shell eval=FALSE} 
#!/bin/bash
STR="library(roxygen2);roxygenize('$1')"
echo $STR
RScript -e $STR
```

### Deploy raadtools package to the local CRAN-like repository

I have the following batch file to build/install/check and finally
write the package source/binaries to a file repository (yes it's
awkwardly hardcoded, but it works and lots of bleeding was involved):

```{shell eval=FALSE}
cd C:\Users\michae_sum\Documents\GIT

rem build for current and ship
rem R CMD build raadtools
rem R CMD INSTALL raadtools --build


RScript -e "p = 'Z:/Transfer/toolbox/RAAD';file.copy('raadtools_0.0-5.tar.gz', contrib.url(p, 'source'), overwrite = TRUE);tools::write_PACKAGES(contrib.url(p, 'source'), type = 'source')"

rem cp raadtools_0.0-5.tar.gz \\aad.gov.au\files\Transfer\toolbox\RAAD\src\contrib
RScript -e "p = 'Z:/Transfer/toolbox/RAAD';file.copy('raadtools_0.0-5.zip', contrib.url(p, getOption('pkgType')), overwrite = TRUE);tools::write_PACKAGES(contrib.url(p, getOption('pkgType')))"

rem build for 2.15 and ship
cd build
C:\inst\R\2.15\R\bin\x64\R CMD INSTALL ../raadtools_0.0-5.tar.gz --build -l .
C:\inst\R\2.15\R\bin\x64\RScript -e "p = 'Z:/Transfer/toolbox/RAAD';file.copy('raadtools_0.0-5.zip', contrib.url(p, getOption('pkgType')), overwrite = TRUE);tools::write_PACKAGES(contrib.url(p, getOption('pkgType')))"

cd ..

```

TODO

tidy up deployment hardcoded values
find a mechanism to build MacOSX packages
host repository on a webserver, not the file share