# Installing the raadtools package

## Setup

To install the _raadtools_ package from within the AAD you
should be able to run the following code from within R. A relatively
recent version of R will be assumed, and definitely no earler than
_R 2.15.0_ can be expected to work.

This is the basic full install and configuration, see below for each step in greater
detail.

```{r eval=FALSE,tidy=FALSE}
setInternet2(TRUE)
install.packages(c("maptools", "raster", "testthat", "ncdf", "rgdal", "rgeos", "knitr"),
                 dependencies = TRUE)
install.packages("raadtools", repos ="file://aad.gov.au/files/Transfer/toolbox/RAAD")
```

Now to load the package, use the usual package loading code here. If
there is a problem with determining the path to the data repository
this will provide a message about that.

```{r eval=FALSE}
library(raadtools)
```

If a message reports that your data repository cannot be found, follow
the instructions to set it and send this package author your details.

## Setup details

The remaining text in this section repeats the above steps
individually, this might be required for customized installations or
development.

If the following step fails you may not have your network settings
setup appropriately, so first run _setInternet2_ and then try to
install _raster_ again. (This option can be set for an R installation
by choosing "Internet2" when R is installed, and this is what you
should do. Otherwise, R is launced with "Rgui.exe" and this can be
invoked with the option to setInternet2 by using "Rgui.exe
--internet2" which can be set on a Windows shortcut under Properties).

```{r eval=FALSE}
setInternet2(TRUE)
install.packages(c("raster", "maptools"))
```

The following packages are not essential to run \textbf{raadtools} but
otherwise some features will not be available. To get the full
functionality please do run the following installations as well.


```{r eval=FALSE}
install.packages("ncdf", "rgdal", "rgeos")
```

Finally, install the latest version of \textbf{raadtools} from the
local AAD repository.

```{r eval=FALSE}
install.packages("raadtools", repos ="file://aad.gov.au/files/Transfer/toolbox/RAAD")
```

## Development of raadtools


