---
title: "Moving raadtools from raster to terra"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Moving raadtools from raster to terra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Every `read*()` function in raadtools now returns a terra `SpatRaster`. It used
to return a raster `RasterLayer`, `RasterBrick` or `RasterStack`.

Nothing else about how you call them has changed. The same function names take
the same arguments and mean the same things, and a call that read a time series
before reads the same time series now.

## The two names to know

`rast()` is terra's `raster()`, `brick()` and `stack()` in one. There is one
class, `SpatRaster`, whether it holds one layer or a thousand.

```{r}
library(terra)

## was: raster("file.tif"), brick("file.nc"), stack(files)
r <- rast("file.tif")
```

`ext()` is terra's `extent()`.

```{r}
## was: extent(r), and extent(xmin, xmax, ymin, ymax)
ext(r)
ext(c(100, 160, -70, -40))
```

Those two carry most of the translation. For everything else, terra's own
comparison table is the reference, and it is the best place to look when a
raster function you use is not mentioned here:

<https://rspatial.github.io/terra/reference/terra-package.html#comparison-with-the-raster-package>

## The ones you are most likely to hit

| raster | terra |
|---|---|
| `raster()`, `brick()`, `stack()` | `rast()` |
| `extent()` | `ext()` |
| `nlayers()` | `nlyr()` |
| `getZ()`, `setZ()` | `time()`, `time()<-` |
| `projection()` | `crs()` |
| `projectRaster()` | `project()` |
| `cellStats(x, "mean")` | `global(x, "mean", na.rm = TRUE)` |
| `rasterToContour()` | `as.contour()` |
| `rasterToPoints()` | `as.data.frame(x, xy = TRUE)` |
| `calc()`, `overlay()` | `app()`, `lapp()` |
| `couldBeLonLat()` | `is.lonlat()` |

`crop()`, `mask()`, `aggregate()`, `resample()`, `writeRaster()`, `plot()` and
`x[[i]]` keep their names and work the way you expect.

One difference that is easy to miss: `values()` returns a **matrix** in terra
where it returned a vector in raster. Ask for a vector explicitly.

```{r}
values(r, mat = FALSE)
```

## What raadtools does about it

The legacy names are **not deprecated**. `readsst()`, `readice()`, `readcurr()`
and the rest are still the front doors, and each one picks a sensible default
and hands back a `SpatRaster`. Use them.

Alongside them are functions naming a specific product, such as
`read_oisst_daily()` or `read_nsidc_cdr_daily()`. Those pin the choice the
legacy name makes for you, so reach for one when you want to be explicit about
which product you are reading rather than take the default.

The first time you call one of the legacy readers in a session, it says so
once:

```
'readsst' now returns a terra SpatRaster, not a Raster* object.
  This call reads through 'read_oisst_daily'.
```

Once per function, per session. It is not repeated on every call, so reading a
long time series in a loop stays quiet. To turn it off entirely:

```{r}
options(raadtools.shim.warn = FALSE)
```

## Two other things worth knowing

`extract()` on a read function still works the same way, and still takes a
data.frame of longitude, latitude and date-time:

```{r}
extract(readsst, xyt)
```

It is now the [xyt](https://github.com/AustralianAntarcticDivision/xyt) package
underneath, which reads only the time slices your points actually need.

`xylim` still accepts everything it always did: a bare numeric of four, a
raster `Extent`, a raster object, or a terra `SpatExtent`. Existing code that
passes one does not need changing.

```{r}
readice("2020-01-15", xylim = ext(c(-1e6, 1e6, -1e6, 1e6)))
```

## If something looks wrong

Please report it. The conversion touched every reader in the package, and the
data sets behind them span decades and several changes of file format each, so
the remaining problems are the ones that only show up against real data and
real questions.

<https://github.com/AustralianAntarcticDivision/raadtools/issues>
